---
title: "purrr"
author: "Jake Riley - 1/14/2022"
output:
  github_document: default # comment out this format to enable moon_reader add-in
  
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
knit:
    (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = ".", output_format = "all")
      }
    )
---

```{r xaringan-themer, eval=TRUE, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_light(
  table_row_even_background_color = "transparent",
  code_inline_color = "black",
  code_inline_background_color = "#f8f8f8",
  header_background_padding = "0px",
  header_background_content_padding_top = "0px",
  link_decoration = "underline"
)

```

```{css css-style, echo=FALSE}
.title-slide {
  vertical-align: bottom !important; 
  text-align: left !important;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

## Agenda

* [slides](https://rjake.github.io/analytics-deep-dive//sessions/purrr/README.html#1) | [code](https://www.github.com/rjake/analytics-deep-dive/blob/main/sessions/purrr/README.md)

* Features of `purrr` - **what & how**

* In practice - **when & why**

* Other goodies

---

## What is `map()`
```{r, eval=TRUE}
library(tidyverse)
```

`map()` comes from `purrr` and most closely relates to `lapply()`
```{r}
lapply(X = 1:2, FUN = rnorm, n = 3)
map(.x = 1:2, .f = rnorm, n = 3)
```

but `purrr` allows twiddle / tilde notation
```{r}
lapply(X = 1:2, FUN = function(x) rnorm(x, n = 3))
map(.x = 1:2,   .f = ~rnorm(.x, n = 3))
#                       ~fn(.x, ...)
```

---

## twiddle / tilde notation

this notation is used throughout the tidyverse
```{r}
# subset data in ggplot
ggplot(mpg, aes(cty, hwy)) +
  geom_point() +
  geom_point(
    data = ~head(.x, 10),
    size = 4, color = "blue"
  )

# convert to km and grab mean
mpg |>
  summarise(
    across(
      .cols = c(cty, hwy), 
      .fns = ~mean(.x * 1.6)
    )
  )
```

---

## There are many Variants

More than one argument

* `map2()` - .x, .y
* `imap()` - use indexes ..1, ..2, etc
* `pmap()` - pass a list or tibble with columns as argument names


Vectors (lapply + unlist)

* `map_chr()`, `map_dbl()`, `map_int()`, etc – return a vector (use `map_dbl()` for dates)

Do an action

* `walk()`, `walk2()`, `iwalk()`, `pwalk()` – doesn't return anything, ex. save a series of plots
* `dplyr` has `group_walk()` and others

Iterate and bind to a data frame

* `map_dfr()`, `map_dfc()` - will bind rows / columns to a table in each iteration

---

## How to replace loops
Loops look like this 

```{r}
for ( some_constant_name    in        some_vector   ) {
    # made here only     not %in%  existing vector/list
  do this
}
```

```{r, eval=TRUE}
for (i in 1:3) {
  print(paste("This is line", i))
}
```


---

## Loops can have several steps
```{r}
for (dept in unique(cohort$department_name)) {
  df <- 
    cohort |>
    filter(department_name == dept)
  
  title <- paste("There were", nrow(df), "visits in", dept)
  
  ggplot(df, aes(month)) +
    geom_bar() +
    labs(title = title)
}
```

---

## Why I don't like loops

Re-running a loop can produce unwanted effects. 

If I accidentally re-run this for-loop without re-running `df <-` it will re-append more data to the data frame and duplcate my data

```{r loop-df-bind-row}
csv_list <- list.files(pattern = "csv$")
df <- data.frame()

for (i in 1:length(csv_list)) {
  x <- read_csv(csv_list[i], na = "NULL")
  df <- bind_rows(df, x)
}

```

---

## How to replace loops with `purrr`

If we take our earlier example:

```{r, ref.label='loop-df-bind-row'}

```


There is a `map_*()` variant that does it all at once, read each csv and bind the rows. 

There is no need for the `csv_list` object and no need to set up an empty data frame `df`


```{r}
df <- 
  map_dfr(
    .x = list.files(pattern = "csv$"),
    .f = read_csv,
    na = "NULL"
  )
```

---

## Break loops into functions + iteration

If you've written loops like this, you've done the heavy lifting

```{r}
for (file in csv_list) {
  x <-
    read_csv(file, na = NULL) |> 
    head(3) |> 
    mutate_if(is.numeric, floor) |>
    mutate(id = row_number())

  df <- bind_rows(df, x)
}
```

You can put the body of the loop into the body of a function. `file` now becomes the argument of the function.

```{r}
simplify_data <- function(file) {
  read_csv(file, na = NULL) |> 
    head(3) |> 
    mutate_if(is.numeric, floor) |>
    mutate(id = row_number())
}
```

---

## How it simplifies

With the new function, you can run try out different inputs to check that it works

```{r}
simplify_data(file = "my_data.csv")
```

Now the loop is simpler
```{r}
for (file in csv_list) {
  x <- simplify_data(file)
  df <- bind_rows(df, x)
}

```

or better yet can be completely eliminated with `map_dfr()`
```{r}
df <-
  map_dfr(
    csv_list,
    simplify_data
  )

```

---

## Errors

`purrr` also lets you catch errors with out the stopping with `possibly()`

```{r}
df <- 
  map_dfr(
    c("123", "my_data.csv", "456"),
    possibly(simplify_data, otherwise = NULL)
  )
```

```
[1] "123"
[1] "my_data.csv"
[1] "456"
Warning messages:
1: In file(file, "rt") : cannot open file '123': No such file or directory
2: In file(file, "rt") : cannot open file '456': No such file or directory

```

I can now troubleshoot one at a time
```{r, eval=FALSE}
simplify_data(file = "123")
simplify_data(file = "my_data.csv")
```

---

class: center, middle

# Putting it all together

---

## Create one row per date

You might use `map2()` to create a vector of dates, this creates a list column `days`
```{r, eval=TRUE}
df <- 
  tibble(
    hospital_admit_date = Sys.Date() + 1:2,
    hospital_discharge_date = hospital_admit_date + 3:4
  ) |> 
  mutate(
    days = 
      map2(
        .x = hospital_admit_date, 
        .y = hospital_discharge_date, 
        .f = seq, 
        by = "1 day"
      )
  ) |> 
  print()

```

---

## Unnest the values

Now use `unnest_longer()` to create one row per date (`unnest()` works also)

```{r, eval=TRUE}
df |> 
  unnest_longer(days)
```

---

## Nested JSON

```{r, eval=TRUE}
df <- 
  tibble(
    all_procedures = '[{
      "cpt": "1234.5",
      "proc_id": "2022",
      "description": "spinal fusion",
      "laterality": "n/a",
      "panel_number": "1",
      "service": "Orthopedic",
      "surgeon_provider_id": "5555",
      "surgeon": "Jake Riley"
    }]'
  ) |> 
  mutate(all_procedures = map(all_procedures, jsonlite::fromJSON)) |> 
  print()
```

---

## Unnested JSON

You can use `unnest()` here but `unnest_wider()` is more explicit
```{r, eval=TRUE}
df |> 
  unnest_wider(all_procedures)
```

---
class: center, middle

# Documents & Dashboards

---

## Iterating with widgets

When using widgets like `plotly` or DT use `walk( )` then `tagList( )`
```{r}
df_list <- 
  mpg |>
  group_split(class) |> # returns a list
  group_walk(DT::datatable)

htmltools::tagList(dt_tables)
```

---

## Iterating with static plots

for `ggplot2` use `walk(..., print)`

```{r, eval=FALSE}
mpg |> 
  head(20) |> 
  group_by(class) |> 
  group_walk(
    ~{
      p <- 
        ggplot(x, aes(cty, hwy)) +
        labs(title = .y[[1]])
      
      print(p)
    }
  )
```

When `group_by()` is used in `group_map()` or `group_walk()` you can find it's group name with `.y[[1]]`. 

See [this SO post](https://stackoverflow.com/a/54755352/4650934) for more info about group titles

---

---
class: center, middle

# Other goodies
---

## `keep()` and `discard()`

retain values with `keep()`
```{r, eval=TRUE}
sentences[1:5] |> 
  keep(str_detect, "^The")

```

remove values with `discard()`
```{r, eval=TRUE}
sentences[1:5] |> 
  discard(str_detect, "^The")
```

---

## `keep()` and `discard()`

you can also use the twiddle with these
```{r, eval=TRUE}
sentences[1:5] |> 
  keep(~nchar(.x) < 40)
```

so much easier to read
```{r}
cohort |> 
  pull(department_name) |>
  discard(str_detect, "BGR") |>
  sort()
```

compare this with 
```{r}
sort(cohort$department_name[!grepl("BGR", cohort$department_name)])

# or
sort(grep("BGR", cohort$department_name, value = TRUE, invert = TRUE))
```

---

## `pluck()`

`pluck()` will grab the nth element. I use it often with `json`. It can use numeric positions or names

```{r, eval=TRUE}
my_list <- 
  list(
    list(x = "abc", name = "Jake"), 
    list(name = "Cody")
  )
```


```{r, eval=TRUE}
# by name
my_list |> 
  map_chr(pluck, "name")


#by index (1 = first)
my_list |> 
  map_chr(pluck, 1)
```

---

## `map2()`

`map2( )` works with 2 inputs of the same length
```{r, eval=TRUE}

map2(
  .x = c("a", "b"),
  .y = c(3, 4), 
  .f = rep    # ex: rep(x = "a", times = 3)
)

```

---

## Simplifying

Since `map()` functions return a list, you need to keep using `map()` functions until it is no longer a list object

```{r}
map2(1:2, 8:9, ~.x/.y) |> # list
  map(mean) |>            # list
  map(round, 3) |>        # list
  map_dbl(pluck)          # vector
```


you can often convert to a vector with the `map_*()` helpers
```{r}
map2(1:2, 8:9, ~.x/.y) |> # list
  map_dbl(mean) |>        # vector
  round(3)                # vector
```
