---
title: "Reproducible Examples in R"
from: markdown+emoji
author: "Jake Riley 3/2/2023"
editor: source
format: 
  revealjs:
    output-file: index.html
    theme: [moon, custom.scss]
    mouse-wheel: true
    scrollable: true
    width: 1200
    height: 750
  gfm:
    output-file: README.md
    
knitr: 
  opts_chunk: 
    echo: true
    message: false
    warning: false
---

```{r setup, include=FALSE}
options(
  tibble.print_min = 6
)

```

## Agenda

* [slides](https://www.github.com/pages/rjake/analytics-deep-dive/sessions/reprex-r/index.html#1) \| [code](https://www.github.com/rjake/analytics-deep-dive/blob/main/sessions/reprex-r/README.md)

* Examples in R

* Live demo

* Recap

* If time / interest... examples in SQL

## Getting help in R

**Pointers**

* use built-in datasets

* `dput()` + `datapasta` to recreate data

* `set.seed()` for random-generating functions

* always include libraries, the reprex package can help

<br>

The following examples will use these libraries

```{r libraries}
library(rocqi)
library(tidyverse)
```


# Why

## Let's look at an example


:::: {.columns}
::: {.column}

Original

```{r}
#| eval: false
#| code-line-numbers: "1-3,15-16"
score_percent <- 
  respondents_clean |> 
  count(group, score_raw) |> 
  mutate(perc = n / sum(n)) |> 
  mutate(
    color = case_when(
      score_raw == 0 ~ "A", 
      between(score_raw, 1, 2) ~ "B", 
      TRUE ~ "C"
    ),
    score_raw = factor(score_raw)
  ) |> 
  arrange(score_raw)

ggplot(score_percent, aes("group", perc, fill = color)) +
  geom_bar(color = "black", position = "stack", stat = "identity")
```

```{r}
#| echo: false
library(lubridate)
library(tidyverse)

tibble(
  score_raw = c(0:4),
  n = c(10, 4:1),
  color = case_when(
    score_raw == 0 ~ "A", 
    score_raw <= 2 ~ "B", 
    TRUE ~ "C"
  )
) |> 
  ggplot(aes("group", n, fill = color)) +
  geom_col(color = "black", position = "fill") +
  theme(aspect.ratio = 1.5)
```
:::
::: {.column}

Reprex

```{r}
#| code-line-numbers: "1,2,4,17,20"
library(lubridate)
library(tidyverse)

score_percent <-
  tibble(
    score_raw = c(0:4),
    n = c(10, 4:1)
  )

score_percent |> 
  mutate(
    color = case_when(
      score_raw == 0 ~ "A", 
      score_raw <= 2 ~ "B", 
      TRUE ~ "C"
    ) |> 
      fct_reorder(n, sum) # <----
  ) |> 
  ggplot(aes("group", n, fill = color)) +
  geom_col(color = "black", position = "fill") +  # <----
  theme(aspect.ratio = 1.5)
```

:::
::::







## Use built-in dates

Use `Sys.time()` and `Sys.Date()` 
```{r dates}
tibble(
  date_requested = Sys.time() + (1:6 * 60),
  surgery_date = Sys.Date() + (1:6 * 7),
  date_dist = surgery_date - as.Date(date_requested)
)
```

I like starting with a `tibble()` instead of `data.frame()` since you can use columns made at the time



## Use built-in `letters`
```{r}
tibble(
  surgeon = rep(LETTERS[1:3], each = 2),
  location = rep(letters[1:3], times = 2),
  op_note = sentences[1:6] # comes from stringr
)
```




## Use `sentences` for regex

This comes from `stringr`
```{r}
tibble(
  note_text = sentences[1:5],
  last_word = str_extract(note_text, "\\w+\\.$") # don't want "."
)
```



## Mock up missing values from scratch

```{r}
tibble(
  x = c(1, NA, 2),
  y = c(NA, "a", NA)
)
```



## Mock up missing values using existing data

```{r}
head(mtcars) |>
  select(1:5) |>
  mutate_all(# make NA if rows are in 2 or 4
    ~ifelse(row_number() %in% c(2, 4), NA, .x)
  )
```



## Use `?fn` to get sharable examples

If you need help with an spc chart, look in the help page

Using `?spc` gives this example:

```{r, eval=FALSE}
spc(
  data = sepsis,
  x = hospital_admit_date,
  y = abx_30_min_ind,
  chart = "p",
  part_dates = "2022-06-01"
)
```



## Use built-in distribution functions

.pull-left[
`rnorm()` gives a normal distribution
```{r, fig.height=5}
hist(
  rnorm(2000, mean = 0.5, sd = 0.1)
)
```
]

.pull-right[
`rbeta()` gives a skewed distribution
```{r, fig.height=5}
hist(
  rbeta(2000, shape1 = 2, shape2 = 10)
)
```
]



## Use built-in distribution functions

You can use this to mock up your data
```{r}
tibble(
  # normal age distribution
  age = rnorm(n = 6, mean = 10, sd = 3),
  # skewed LOS distribution up to 2 weeks
  los_days = rbeta(n = 6, shape1 = 2, shape2 = 10) * 14
            # ------------- 0 to 1 ------------ x 14 days
)
```



## Use `set.seed()` for reproducible randomness

Without a "seed", random number functions like `rnorm()` will give different results each time
```{r}
rnorm(5)
rnorm(5)
```



## Use `set.seed()` for reproducible randomness

Using `set.seed()` will make sure that the random number generator starts at the same point each time. Note: it needs to be right before random fn or in the right order
```{r}
set.seed(2022) 
rnorm(5)

set.seed(2022) 
rnorm(5)
```



## The `reprex` package

Copy to code to your clipboard then use `reprex::reprex_slack()` in console or highlight code in source pane and use `reprex` add-in
```{r, eval=FALSE}
mpg |>
  mutate(x = Sys.Date)
```

```r
mpg |>
  mutate(x = Sys.Date)
#> Error in mutate(mpg, x = Sys.Date): could not find function "mutate"
```



## `dput()`

Use `dput()` to "export" the structure of your data.
```{r}
your_data <- head(ToothGrowth, 3)
dput(your_data)
```



## `structure()`

Assign the results of `dput()` to a new object

```{r}
my_data <-
  structure(
    list(
      len = c(4.2, 11.5, 7.3),
      supp = structure(c(2L, 2L, 2L), .Label = c("OJ", "VC"), class = "factor"),
      dose = c(0.5, 0.5, 0.5)
    ),
    row.names = c(NA, 3L),
    class = "data.frame"
  )

identical(your_data, my_data)
```



## `clipr` and `datapasta`

Copy data to your clipboard

```{r, eval=FALSE}
your_data |> clipr::write_clip() 
```

Use add-ins: `datapasta` > `paste as tribble`. Also works with data on your clipboard from excel

```{r}
tibble::tribble(
  ~len, ~supp, ~dose,
  4.2, "VC", 0.5,
  11.5, "VC", 0.5,
  7.3, "VC", 0.5
)
```



## Keep things minimal

Just select the colums you need and a minimal # of rows
```{r}
#   row  col
mpg[1:3, 1:2]
```

A row of 1 is often enough
```{r}
head(mpg, 1) |> #dput()
  mutate_if(is.character, toupper)
```




## Built-in data 
* see all data sets with
    * `data()`
    * `data(package = "rocqi")`

* base R
    * `iris`
    * `ToothGrowth`
    * `quakes` - mapping

* [`ggplot2`](https://ggplot2.tidyverse.org/reference/index.html#section-data) - all 3 good for plotting, mix of categorical & numeric
    * `mpg`
    * `diamonds`
    * `msleep`
    * `map_data` - map polygons ([see vignette](https://ggplot2.tidyverse.org/reference/map_data.html))



## Built-in data 

* [`tidyr`](https://tidyr.tidyverse.org/reference/index.html#data)
    * `us_rent_income` - pivoting
    * `population` - longitudinal

* [`dplyr`](https://dplyr.tidyverse.org/reference/index.html#section-data)
    * `storms` - longitudinal
    * `starwars` - lists, missingness, comma separated values



## Built-in data 

  * `sf`
      * `st_read(system.file("shape/nc.shp", package="sf"))` - comes with install


  * [`rocqi`](https://github.research.chop.edu/pages/analytics/rocqi/reference/index.html#section-data-sets)
      * `ed_fractures`
      * `surgeries`

## This app

https://rjake.shinyapps.io/Find-Good-Data/


## R Takeaways
**Always**
* give all code required to reproduce the problem
* remove extraneous code
* include expected output if possible, use pictures, arrows, etc.

**For R**
* use built-in datasets
* `dput()` + `datapasta` to recreate data
* `set.seed()` for random-generating functions
* always include libraries, the reprex package can help


# Live Demo

# SQL

## Getting help in SQL

**Pointers**

* use `select` without a `from` statement

* mock up CTEs with `select` + `union`

* use reserved words like `current_date` to mock up data types

* write a small sample to `qmr_dev` if all else fails

```{r sql-knitr, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE
)
```



## can use select alone, no 'from' statement
```{sql reprex-sql}
select 123
```



## Use built-in date values

`current_date` and `current_timestamp` are built in
```{sql}
select
    current_date - 2 as encounter_date,
    current_timestamp as surgery_time, -- also now()
    surgery_time::date - encounter_date as n_days,
    datetime(timezone(now(),  'UTC', 'America/New_York')) as time_utc    
```



## Mock-up note_text

```{sql}
select
    'I want the second word' as note_text,
    regexp_extract(note_text, '\w+', 1, 2) as second_word
```



## Set-up a small data set with a `union` CTE
```{sql}

with 
    clinic_list as (
              select '555-123-4567' as phone_number
        union select '555.123.4567'
        union select '(555)123-4567'
        union select '1-555-123-4567' 
        union select '(555)123-4567 ext. 9' 
    )
select 
    phone_number,
    regexp_replace(
        phone_number,
        '^1?[^\d]*(\d{3})[^\d]*(\d{3})[^\d]*(\d{4}).*',
        '\1-\2-\3'
    ) as clean_number
from clinic_list
```



## Put a subset of the data in `QMR_DEV`

Put subset in qmr_dev for someone to help you if it does require a few CTEs to get to your output add a small helper `select` at the end of the CTE

.pull-left[
```{sql}
*--create table abc as 
with
    x as (
       select 
       from
*    ) --select * from x;

    , y as (
       select 
       from
    )

select *
from 
   ...

```
]

.pull-right[
Remember to drop anything you don't need long-term
```{sql}
drop table abc if exists;
```
]


## SQL Takeaways
* use `select` without a `from` statement
* mock up CTEs with `select` + `union`
* use reserved words like `current_date` to mock up data types
* write a small sample to `qmr_dev` if all else fails
