---
title: "Reproducible Examples in R"
from: markdown+emoji
author: "Jake Riley 3/2/2023"
editor: source
format: 
  revealjs:
    output-file: index.html
    theme: [moon, custom.scss]
    mouse-wheel: true
    scrollable: true
    width: 1200
    height: 750
    slide-number: c/t
    controls: true
    df-print: kable
    #navigation-mode: grid
  gfm:
    output-file: README.md
    
knitr: 
  opts_chunk: 
    echo: true
    message: false
    warning: false
---

```{r setup, include=FALSE}
options(
  tibble.print_min = 6
)

```

## Agenda

* [slides](https://rjake.github.io/analytics-deep-dive/sessions/reprex-r) \| [code](https://www.github.com/rjake/analytics-deep-dive/blob/main/sessions/reprex-r/README.md)

* What is a reproducible example

* How to mock up data

* Built in data sets

* Recap

* If time / interest... examples in SQL

<br>

We will use these libraries

```{r libraries}
library(tidyverse)
library(rocqi) # internal to CHOP, see https://github.research.chop.edu/pages/CQI/chopran/
```

# Reproducible Examples

`reprex` is shorthand for a `rep`roducible `ex`ample
<br>
It should be as little code as required for other people to run your code


## Let's look at an example


:::: {.columns}
::: {.column}

Original

```{r}
#| eval: false
#| code-line-numbers: "1-4,15-16"
score_percent <- 
  respondents_clean |> 
  count(group, score_raw) |> 
  mutate(perc = n / sum(n)) |> 
  mutate(
    color = case_when(
      score_raw == 0 ~ "A", 
      between(score_raw, 1, 2) ~ "B", 
      TRUE ~ "C"
    ),
    score_raw = factor(score_raw)
  ) |> 
  arrange(score_raw)

ggplot(score_percent, aes("group", perc, fill = color)) +
  geom_bar(color = "black", position = "stack", stat = "identity")
```

```{r}
#| echo: false
library(lubridate)
library(tidyverse)

tibble(
  score_raw = c(0:4),
  n = c(10, 4:1),
  color = case_when(
    score_raw == 0 ~ "A", 
    score_raw <= 2 ~ "B", 
    TRUE ~ "C"
  )
) |> 
  ggplot(aes("group", n, fill = color)) +
  geom_col(color = "black", position = "fill") +
  theme(aspect.ratio = 1.5)
```
:::
::: {.column}

Reprex

```{r}
#| code-line-numbers: "1,2,4,6,7,17,20"
library(lubridate)
library(tidyverse)

score_percent <-
  tibble(
    score_raw = c(0:4),
    n = c(10, 4:1)
  )

score_percent |> 
  mutate(
    color = case_when(
      score_raw == 0 ~ "A", 
      score_raw <= 2 ~ "B", 
      TRUE ~ "C"
    ) |> 
      fct_reorder(n, sum) # <----
  ) |> 
  ggplot(aes("group", n, fill = color)) +
  geom_col(color = "black", position = "fill") +  # <----
  theme(aspect.ratio = 1.5)
```

:::
::::



## Getting help in R

**Pointers**

* use built-in datasets

* `dput()` + `datapasta` to recreate data

* `set.seed()` for random-generating functions

* always include libraries, the `reprex` package can help




# Built-in Data

There are data sets and constants in R that can help you mock up your code

## Use built-in dates

Use `Sys.time()` and `Sys.Date()` 
```{r dates}
tibble(
  date_requested = Sys.time() + (1:6 * 60),
  surgery_date = Sys.Date() + (1:6 * 7),
  date_dist = surgery_date - as.Date(date_requested)
)
```

I like starting with a `tibble()` instead of `data.frame()` since you can use columns made at the time



## Use built-in `letters`

Note `rep(..., each)` vs `rep(..., times)`

```{r}
tibble(
  surgeon = rep(LETTERS[1:3], each = 2),
  location = rep(letters[1:3], times = 2),
  op_note = sentences[1:6] # comes from stringr
)
```




## Use `sentences` for regex

This comes from `stringr`
```{r}
tibble(
  note_text = sentences[1:5],
  last_word = str_extract(note_text, "\\w+\\.$") # don't want "."
)
```



## Mock up missing values from scratch

```{r}
tibble(
  x = c(1, NA, 2),
  y = c(NA, "a", NA)
)
```



## Mock up missing values using existing data

```{r}
head(mtcars) |>
  select(1:5) |>
  mutate_all(# make NA if rows are in 2 or 4
    ~ifelse(row_number() %in% c(2, 4), NA, .x)
  )
```



## Use `?fn` to get sharable examples

If you need help with an spc chart, look in the help page

Using `?spc` gives this example:

```{r, eval=FALSE}
spc(
  data = sepsis,
  x = hospital_admit_date,
  y = abx_30_min_ind,
  chart = "p",
  part_dates = "2022-06-01"
)
```


# Random numbers

## Use built-in distribution functions

:::: {.columns}
::: {.column}
`rnorm()` gives a normal distribution
```{r, fig.height=5}

hist(
  rnorm(2000, mean = 12, sd = 3)
)
```
:::

::: {.column}
`rbeta()` gives a skewed distribution
```{r, fig.height=5}
hist(
  rbeta(2000, shape1 = 2, shape2 = 10)
)
```
:::
::::



## Use built-in distribution functions

You can use this to mock up your data
```{r}
tibble(
  # normal age distribution
  age = rnorm(n = 6, mean = 12, sd = 3),
  # skewed LOS distribution up to 2 weeks
  los_days = rbeta(n = 6, shape1 = 2, shape2 = 10) * 14
            # ------------- 0 to 1 ------------ x 14 days
)
```



## Use `set.seed()` for reproducible randomness

Without a "seed", random number functions will give different results each time
```{r}
rnorm(5)
rnorm(5)
```

<br><br>
Using `set.seed()` will make sure that the random number generator starts at the same point each time. Note: it needs to be right before random fn or in the right order
```{r}
set.seed(2023) 
rnorm(5)

set.seed(2023) 
rnorm(5)
```


# The `reprex` package

Developed by Jenny Bryan and widely used

## It's easy to use

Copy code to your clipboard then use `reprex::reprex_selection()` in console 
<br> or highlight code in source pane and use `reprex` add-in
```{r, eval=FALSE}
mtcars |>
  mutate(x = Sys.Date)
```


    ```r
    mpg |>
      mutate(x = Sys.Date)
    #> Error in mutate(mpg, x = Sys.Date): could not find function "mutate"
    ```

Try again...
```{r, eval=FALSE}
library(tidyverse)

mtcars |>
  mutate(x = Sys.Date)
```


    ```r
    library(tidyverse)
    
    mpg |>
      mutate(x = Sys.Date)
    #> Error in `mutate()`:
    #> ℹ In argument: `x = Sys.Date`.
    #> Caused by error:
    #> ! `x` must be a vector, not a function.
    ```
One more time...
```{r, eval=FALSE}
library(tidyverse)

mtcars |>
  mutate(x = Sys.Date())
```

:tada: :tada: :tada:

    ``` r
    library(tidyverse)
    
    mtcars |>
      mutate(x = Sys.Date())
    #>                      mpg cyl  disp  hp drat    wt  qsec vs am gear carb          x
    #> Mazda RX4           21.0   6 160.0 110 3.90 2.620 16.46  0  1    4    4 2023-03-02
    #> Mazda RX4 Wag       21.0   6 160.0 110 3.90 2.875 17.02  0  1    4    4 2023-03-02 
    #> Datsun 710          22.8   4 108.0  93 3.85 2.320 18.61  1  1    4    1 2023-03-02
    #> Hornet 4 Drive      21.4   6 258.0 110 3.08 3.215 19.44  1  0    3    1 2023-03-02
    #> Hornet Sportabout   18.7   8 360.0 175 3.15 3.440 17.02  0  0    3    2 2023-03-02
    ```


# Actually send your data

## `dput()` & `structure()`

Use `dput()` to "export" the structure of your data.
```{r}
your_data <- head(ToothGrowth, 3) |> as_tibble()
dput(your_data)
```


Assign the results of `dput()` to a new object

```{r}
my_data <-
  structure(
    list(
      len = c(4.2, 11.5, 7.3),
      supp = structure(c(2L, 2L, 2L), .Label = c("OJ", "VC"), class = "factor"),
      dose = c(0.5, 0.5, 0.5)
    ),
    row.names = c(NA, 3L),
    class = c("tbl_df", "tbl", "data.frame")
  )

identical(your_data, my_data)
```

## `datapasta`

Make a reproducible example with this or the add-in: `datapasta` > `fiddle_selection`

```{r, eval=FALSE}
your_data |> datapasta::dpasta()
```

```{r}
your_data <-
  tibble::tribble(
    ~len, ~supp, ~dose,
     4.2,  "VC",   0.5,
    11.5,  "VC",   0.5,
     7.3,  "VC",   0.5
  )
```

For data copied from excel, copy to your clipboard and use the add-in: `datapasta` > `paste as tribble`.


## Keep things minimal

Just select the colums you need and a minimal # of rows
```{r}
starwars |> 
  select(1:3) |> 
  head(2)
```

A row of 1 is often enough
```{r}
head(mpg, 1) |> # use dput() here
  mutate_if(is.character, toupper)
```



```{r sql-knitr, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE
)
```

# Built-in data 

see all data sets with
```{r}
data()
data(package = "dplyr")
```

## Some good ones to know
base R
```{r}
iris
mtcars
airquality
Orange
ToothGrowth
quakes # mapping
```


## How is this helpful?

:::: {.columns}
::: {.column}
Rename columns for plots
```{r}
#| eval: true
#| code-line-numbers: "2,5,6"
df <- 
  ggplot2::economics |>
  head(3) |> 
  select(
    recorded_date = date,
    lab_value = pce
  )
```
:::

::: {.column}
```{r}
#| eval: true
#| echo: false
df
```

:::
::::

Mix and match pieces
```{r}
#| eval: true
#| code-line-numbers: "2-5"
tibble(
  hospital_admit_date = economics$date[1:3],
  losandays = economics$uempmed[1:3],
  name = starwars$name[1:3],
  note_text = stringr::sentences[1:3]
)
```


## Others

:::: {.columns}
::: {.column}
[ggplot2](https://ggplot2.tidyverse.org/reference/index.html#section-data) 
```{r}
mpg
diamonds
msleep
map_data("state")
```

[tidyr](https://tidyr.tidyverse.org/reference/index.html#data)
```{r}
us_rent_income # pivoting
population # longitudinal
```

[dplyr](https://dplyr.tidyverse.org/reference/index.html#section-data)
```{r}
storms # longitudinal
starwars # lists, NA values, comma separated
```
:::

::: {.column}
[sf](https://r-spatial.github.io/sf/articles/sf3.html)
```{r}
# some files comes with install
system.file("shape/nc.shp", package="sf") |>
  st_read()
```

[rocqi](https://github.research.chop.edu/pages/analytics/rocqi/reference/index.html#section-data-sets)

```{r}
ed_fractures
surgeries
```

:::
::::

<br>

Also this app

[https://rjake.shinyapps.io/Find-Good-Data](https://rjake.shinyapps.io/Find-Good-Data/)

## Takeaways

**Always**

* give all code required to reproduce the problem
* remove extraneous code
* include expected output if possible, use pictures, arrows, etc.
* use built-in datasets
* `dput()` + `datapasta` to recreate data
* `set.seed()` for random-generating functions
* always include libraries, the reprex package can help


# SQL

## Use `select` without a `from` statement

```{sql reprex-sql}
select 123
```



## Use built-in date values

`current_date` and `current_timestamp` are built in
```{sql}
select
    current_date - 2 as encounter_date,
    current_timestamp as surgery_time, -- also now()
    surgery_time::date - encounter_date as n_days,
    datetime(timezone(now(),  'UTC', 'America/New_York')) as time_utc    
```



## Mock-up note text

```{sql}
select
    'I want the second word' as note_text,
    regexp_extract(note_text, '\w+', 1, 2) as second_word
```



## Set-up a small data set with a `union` CTE
```{sql}

with 
    clinic_list as (
              select '555-123-4567' as phone_number
        union select '555.123.4567'
        union select '(555)123-4567'
        union select '1-555-123-4567' 
        union select '(555)123-4567 ext. 9' 
    )
select 
    phone_number,
    regexp_replace(
        phone_number,
        '^1?[^\d]*(\d{3})[^\d]*(\d{3})[^\d]*(\d{4}).*',
        '\1-\2-\3'
    ) as clean_number
from clinic_list
```



## Put a subset of the data in `QMR_DEV`

Put subset in qmr_dev for someone to help you if it does require a few CTEs to get to your output add a small helper `select` at the end of the CTE

:::: {.columns}
::: {.column}
```{sql}
--create table abc as 
with
x as (
   select 
   from
) --select * from x;

, y as (
   select 
   from
)

select *
from 
   ...

```
:::

::: {.column}
Remember to drop anything you don't need long-term
```{sql}
drop table abc if exists;
```
:::
::::

# Live Demo

[https://github.com/rjake/analytics-deep-dive/blob/main/sessions/reprex-r/sql-reprex.sql](https://github.com/rjake/analytics-deep-dive/blob/main/sessions/reprex-r/sql-reprex.sql)


## SQL Takeaways

* use `select` without a `from` statement
* mock up CTEs with `select` + `union`
* use reserved words like `current_date` to mock up data types
* write a small sample to `qmr_dev` if all else fails

```{r quarto}
#| include: false
#| eval: false
rstudioapi::getSourceEditorContext()$path |>  
  quarto::quarto_render(output_format = "all")

# CLI > quarto render reprex-r.qmd --to gfm,revealjs
```

